---
title: "Week 1 - Homework"
format: 
  html: default
categories: 
  - week01
  - homework
editor_options: 
  chunk_output_type: console
---

Please directly write your code in a Quarto document. (File -> New File -> Quarto Document -> then click "create")

1. Please import the data file "visceral_fat.csv" and upload it into your folder of choice, then import the data and call it `visceral_fat`. 

You can copy paste the commands R provided by using the click-button way into your Quarto report code chuck. Then you can use this data in your report.

```{r}

```

2. Run the function `summary(visceral_fat)`, describe what you obtained.

```{r}

```

3. Draw a histogram for the diastolic blood pressure `visceral_fat[["diastolic_bp__mmhg"]]`, describe what you observed.

```{r}

```

4. After a double check with a clinician, the 27th value of the diastolic blood pressure was not correct recorded. You can ignore this value by writing `visceral_fat[["diastolic_bp__mmhg"]][-27]`. Redo a histogram after ignoring this outlier, describe what you observed.

```{r}

```

5. Calculate the mean (`mean()`) and the median (`median()`) of the diastolic blood pressure before and after ignoring the 27th value, *i.e.*, `visceral_fat[["diastolic_bp__mmhg"]]` and `visceral_fat[["diastolic_bp__mmhg"]][-27]`, describe what you observed.

```{r}


```

6. Click "Render" to check your report.
