---
title: "Week 5 - Hands-On Examples"
format: 
  html: default
  # pdf: default
date: "2025-02-04"
date-format: " "
categories: 
  - week05
  - exercise
editor_options: 
  chunk_output_type: console
---

The R script is available here:
[link](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/week05/r_w05_exos.R)

## Goals

- Install new R packages
- Create a basic plot with ggplot2
- Create a volcano plot with ggplot2

## Install New Packages

We will use a couple of additional R packages for this training in sessions 8 and 9:

- {[tidyr](https://tidyr.tidyverse.org)}: provides functions that help you get to tidy data
- {[dplyr](https://dplyr.tidyverse.org)}: for data manipulation
- {[tibble](https://tibble.tidyverse.org/)}: a modern re-imagining of the data frame
- {[stringr](https://stringr.tidyverse.org/)}: provides functions designed to make working with strings as easy as possible
- {[readr](https://readr.tidyverse.org)}: provides fast and friendly way to read rectangular data 

Instead of installing them one by one, we can install the {[tidyverse](https://www.tidyverse.org)} package (a collection of previous packages designed for data science).

1. Install the {`tidyverse`} package.

```{r}
# install.packages("tidyverse") # you only need to install it once
```

2. Check if you can find `{dplyr}` and `{tidyr}` in the "*Packages*" panel in RStudio.

3. Please use the following code to create a data frame of all installed packages in your system.
Show a couple of first lines of the data frame.

```{r}
my_pkgs <- as.data.frame(installed.packages()[, c(1, 3:4)])
head(my_pkgs)
```

4. According to `my_pkgs`, how many packages are installed in your system?

```{r}
nrow(my_pkgs) # the number can be different
```

5. Extract the rows for the "ggplot2", "tidyr", "dplyr" packages from `my_pkgs`.

```{r}
# extract by rownames
my_pkgs[c("ggplot2", "tidyr", "dplyr"), ]

## or
# extract by filtering the names in the "Package" column
my_pkgs[my_pkgs$Package %in% c("ggplot2", "tidyr", "dplyr"), ]
```


## Create a Basic ggplot

We'll use the built-in data set `iris`.

0. Load the {`ggplot2`} package.

```{r}
library(ggplot2)
```

1. Create a violin for the petal length by species.

```{r}
p_base <- ggplot(iris, aes(x = Species, y = Petal.Length)) +
  geom_violin()

p_base
```

2. Modify labels:

- Remove x-axis title
- Change y-axis title to "Petal Length (cm)"
- Add a plot title "Distribution of iris' petal length"
- Add a subtitle "The data were collected by Anderson, Edgar (1935)"

```{r}
p_base <- p_base + labs(
  x = NULL,
  y = "Petal Length (cm)",
  title = "Distribution of iris' petal length",
  subtitle = "The data were collected by Anderson, Edgar (1935)"
)
p_base
```


3. Add a layer of boxplot on the violin plot.

- Color the boxes by species.
- Use the `alpha` parameter to let boxes be semi-transparent.

```{r}
p_base <- p_base +
  geom_boxplot(
    aes(color = Species),
    alpha = 0.5
  )
p_base
```

4. Change the theme to `theme_minimal()`. 

```{r}
p_base <- p_base + theme_minimal()
p_base
```

5. With the `theme()` function:

- Increase the font size of x-axis' text.
- Move the legend to the top of the plot.

```{r}
p_base <- p_base + theme(
  axis.text.x = element_text(size = rel(2)),
  legend.position = "top"
)
p_base
```

## Create a Volcanot Plot with ggplot2

We will use the same diffenrential expression (DE) analysis result as in the hands-on example of session 4, which is also available [here](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/exos_data/toy_DEanalysis.csv).

Reminder: the results were obtained by comparing SET1 samples to WT samples using
the data from the [`read-counts.csv`](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/exos_data/read-counts.csv) file.


1. Import the results file in R and name it `de_res`.

```{r}
de_res <- read.csv(
  file = "../exos_data/toy_DEanalysis.csv",  # replace the path with your own
  header = TRUE
)
```

2. Create a new column (`gene_signif`) which indicates whether the adjusted p-value is smaller than 0.05.

```{r}
de_res$gene_signif <- de_res$padj < 0.05
```

:::{.callout-tip}
**Volcano Plot**

A volcano plot is a type of scatter plot commonly used in biological data analysis, particularly for differential expression (DE) analysis. It visually combines two key aspects:

- **Significance** ($-log_{10}(\text{p-value})$): How confident we are that a result is not due to chance.
- **Effect size** ($log_{2}(\text{Fold Change})$): The magnitude of change between conditions.

Volcano plot highlights genes (or features) that are both statistically significant and biologically meaningful.
:::

3. Create a volcano plot using the DE results `de_res`.

- Let the log2 fold change be on the x-axis.
- Transform the adjusted p-value to `-log10()`, then let them be on the y-axis.
- Color the points by whether the adjusted p-value is smaller than 0.05 (*i.e.*, the `gene_signif` column).
- Use the shape 18 for the points.
- Increase the points size to 3.
- Use the `alpha` parameter to make points transparent.
- Add a title for the plot, show number of analysed genes in the subtitle.
- Use the `theme_bw()` for the plot.
- Hide the legend.

```{r}
ggplot(
  data = de_res,
  mapping = aes(x = log2FoldChange, y = -log10(padj))
) +
  geom_point(
    aes(color = gene_signif),
    shape = 18,
    size = 3,
    alpha = 0.5
  ) +
  labs(
    title = "Volcano plot for DE analysis",
    subtitle = paste("Total genes analysed:", nrow(de_res))
  ) +
  theme_bw() +
  theme(legend.position = "none")
```


---

#### Good job! ðŸ‘ðŸ‘ You've taken your first big steps into ggplot2, keep it up! 


