---
title: "Week 2 - Homework"
format: 
  html: default
categories: 
  - week02
  - homework
editor_options: 
  chunk_output_type: console
---

Write your code directly in a Quarto document.

To create a Quarto document: go to **File** -> **New File** -> **Quarto Document**, then click **Create**.

Use following code import the file "read-counts.csv" 
(you have already downloaded it in the hand-on examples of the week01),
name the imported data `expr_data`.

```{r}
expr_data <- read.table(
  file = "read-counts.csv",  # replace the path by the yours
  header = TRUE, sep = ",", row.names = 1
)
```

1. Is the imported data a matrix?

```{r}

```

2. Transform it to a matrix by using `as.matrix()`

```{r}

```

3. What is the dimension of the matrix?

```{r}

```

4. What are the column names and row names?

```{r}

```


5. Extract all genes' expression for the 10 WT samples (WT.1, WT.2, ..., WT.10), store it in a variable called `expr_wt`.
Show the first rows by using `head()` function.

```{r}

```

6. What is the average expression level across all WT samples for each gene? Store the results in a variable called `avg_wt`.

```{r}

```

7. Draw a histogram to check the distribution of the average gene expression level for WT samples.

```{r}

```

8. Repeat question 5 to 7 for the 10 SET1 samples (SET1.1, SET1.2, ..., SET1.10).

```{r}

```

9. What is the absolute difference between the average expression of WT and SET1 samples for each gene? 
Which gene shows the biggest difference?

Hints:

- use `abs()` to get the absolute value. (`?abs`)
- the `max()` function helps to find the maximum value from a vector. (`?max`)
- the `which.max()` function helps to find the index of the maximum value from a vector. (`?which.max`)

```{r}

```


10. Click "Render" to generate your Quarto report.
