---
title: "Data Foundations:<br>Vectors and Matrices"
subtitle: "IOC-R Week2"
format:
  revealjs: 
    toc: true
    toc-title: "Overview"
    toc-depth: 1
    slide-number: true
    preview-links: auto
    logo: ../../images/logoInforBio_fond_blanc_small.png
    css: ../custom.css
echo: true
categories: 
  - week02
  - slides
---

# Data Types in R

## Data Types {.smaller}

Examples: `1`, `2.5`, `A`, `InforBio`, `I love R`, `TRUE`, `FALSE`

What are their data type in the daily life?

. . .

And in R?

- Numeric: `1`, `2.5`
- Integer: `1`
- Character: `"A"`, `"InforBio"`, `"I love R"`
- Logical: `TRUE`, `FALSE`
- Factor: for categorical data

## Where Do We Store Data?  {.smaller}

- A **variable** is like a box where you store data.
- Each variable has a **name** and **content** (one or multiple values).
- A variable is created at the moment you assign a value to it. Use `<-` for assignment.

```{r}
x <- 1 # put a space on each side of the assignment sign
x # type variable name to print its value(s)

char_name <- "InforBio"
char_name

passed <- TRUE
passed
```

Check the "Environment" pane or type `ls()` in the console, are variables you just created there?

::: {.notes}
Variables are containers for data.
A box labeled "x" containing the number 1.
A box labeled "char_name" containing the platform name "InforBio".
:::

## Variable Naming Convention {.smaller}

::: columns
::: {.column}
- Be descriptive and keep it short
- Use snake case (only `[a-z]` and `_`)
- Avoid special characters (such as `!`, `#`, ) and spaces
- Do not start a name with numbers
- Avoid reserved keywords in R (*e.g.*, function, if, TRUE)
- Do not overwrite built-in functions (*e.g.*, mean, sd)
:::

::: {.column .fragment}

Which are validated names?

``` {.r code-line-numbers="1-16|5,9-11"}
foo
test
var
var2
exam_results
a_variable_with_a_name_super_long
count.matrix 

day_1
day_one
day1
first_day_of_the_month
DayOne
dayOne
DAYONE
DAYone
```

:::
:::

::: {.content-smaller .fragment}
How to know if a variable name was already used?

- Type `help(reserved)` to check reserved words in R.
- Check in "Environment" pane.
- Type the first letter of a name and press the **Tab** key to trigger autocompletion
:::

::: {.notes}
Follow the same rules when you naming your R project folder.
If really needs a long name, use abbreviation of it with a comment line at the begin to explain what it stands for.
:::

## Numeric, Integer, Double  {.smaller}

```{r}
my_var <- 10
my_var
class(my_var)
is.numeric(my_var)
```

. . .

```{r}
class(10L) # "L" declares it's an integer
typeof(10L)
```

. . .

```{r}
class(1.5)
typeof(1.5)
```

## Character {.smaller}

R stores text (strings) as character. Use quotation marks to indicate a value is character.

```{r}
# create a vector of characters
x <- c("I", "love", "R")
x

class(x)
is.character(x)
```

. . .

```{r}
#| error: true

x <- "1"
y <- "2"
x + y
```

. . .

```{r}
as.numeric(x) + as.numeric(y)  # convert to numeric
```

. . .

```{r}
as.character(1) # convert back to character
```

## Logical (1) {.smaller}

- Only two possible values for logical data: `TRUE` or `FALSE`.

- Can be written as `T` or `F`, but never in other formats (*e.g.*: `True`, `true`, *etc.*)

```{r}
#| error: true

is.logical(TRUE)
is.logical(T)
is.logical(True)
```

- Can be obtained from logical statements, *e.g.*:

```{r}
2 > 1
```

## Logical (2) {.smaller}

- Convert to other types:

```{r}
as.numeric(TRUE)
as.numeric(FALSE)

vec_logical <- c(TRUE, TRUE, FALSE)
sum(vec_logical)

```

. . .

```{r}
as.character(TRUE)
as.character(FALSE)
```

. . .

```{r}
as.logical(1)
as.logical(-1)
as.logical(0)
```

## Factor {.smaller}

Use factor when you want R treats your data as categorical data.

```{r}
grades <- c("one", "two", "one", "three", "two")
grades

factor(grades)
as.factor(grades)
```

. . .

Specify the levels:

```{r}
grades_fac <- factor(grades, levels = c("one", "two", "three")) # specify the order
grades_fac
```

. . .

Convert to numeric values:

```{r}
as.numeric(grades_fac)
```


# Data Strucutres

## From Single Value to Multiple Values {.smaller}

When we store multiple values, we need a structure.

```{r}
gene1 <- 10
gene2 <- 12
gene3 <- 9
```

How to put `gene1`, `gene2` and `gene3` together?

. . .

<br>
R provides 4 data structures to store multiple values:

|                      | 1 dimension | 2 dimensions (row/column) |
|----------------------|:-----------:|:------------:|
| **Same** data type       | vector      | matrix       |
| **Different** data types | list        | data frame   |


# Vectors

The simplest data structure in R, for **one dimension** data of the **same type**.

## Vector Creation (1) {.smaller}

Use the function `c()` to create a vector and use `,` to separate elements.

```{r}
c(10, 12, 9) # Numeric vector
c(gene1, gene2, gene3)

gene_expr <- c(gene1, gene2, gene3)
gene_expr
c(gene_expr, 18)

# other ways for numeric vector
1:10
seq(10)
```

. . .

```{r}
c("gene1", "gene2", "gene3") # Character vector

c(TRUE, FALSE, FALSE, TRUE, TRUE) # Logical vector
```


## Vector Creation (2) {.smaller}

When you combine different data types ...

```{r}
c(10, TRUE)
c(10, "gene1")
c(TRUE, "gene1")
c(10, "gene1", TRUE)
```

. . .

<br>
R follows a hierarchy of data types for coercion:

logical (**least** flexible) → numeric → character (**most** flexible)

::: {.notes}
R always picks the "most flexible" type that can represent all the values without losing information.
:::


## Vector Indexing (1) {.smaller}

::: columns
::: {.column}

- Use `[`idx`]` to access element(s). 

*Notes*: The index starts from **1**.

```{r}
gene_expr
gene_expr[1]   # 1st element
gene_expr[c(2, 3)] # elements 2 and 3
gene_expr[2:3] # elements 2 and 3
gene_expr[-1]  # remove 1st element
```

:::

::: {.column .fragment}

- Modify element(s).

```{r}
gene_expr[1] <- 100
gene_expr

gene_expr[2:3] <- 8
gene_expr

gene_expr[2:3] <- c(0, 20)
gene_expr
```

:::
:::

## Vector Indexing (2) {.smaller}

::: columns
::: {.column}

- Use `[`name`]` to access element(s) if the vector is named. 

```{r}
# name elements of the vector
names(gene_expr) <- paste0("gene", 1:3)
gene_expr

gene_expr["gene1"] # the element named "gene1"
gene_expr[c("gene1", "gene2")] # extract "gene1" and "gene2"
```

::: {.fragment}

```{r}
#| error: true

gene_expr[-c("gene1", "gene2")] # error
```

:::

:::

::: {.column .fragment}

- Modify element(s).

```{r}
gene_expr["gene1"] <- 2
gene_expr
```


::: {.fragment .content-smaller}
Your turns:

- modify the expression value of "gene1" and "gene2" to 5
- change the expression value of "gene1" to 0 and "gene2" to 16

:::

::: {.fragment}

```{r}
gene_expr[c("gene1", "gene2")] <- 5
gene_expr

gene_expr[c("gene1", "gene2")] <- c(0, 16)
gene_expr
```

:::

:::
:::

## Vector Indexing (3) {.smaller}

::: columns
::: {.column width="60%"}

- Use a logical vector for indexing.

```{r}
num_vec <- c(1, 2, 5, 4)
num_vec
logical_vec <- c(TRUE, TRUE, FALSE, FALSE)
logical_vec

num_vec[logical_vec]
```


```{r}
# create logical vect using comparison operator
num_vec < 3

# then use it to extract values from the numeric vector
num_vec[num_vec < 3]
```

:::

::: {.column width="40%" .fragment}
Your turns:

R use the `==` to test equality. Use this operator to extract the value `5` from the `num_vec`.

```{r}
num_vec[num_vec == 5]
```

:::
:::

## Vector Operations {.smaller}

::: columns
::: {.column}
```{r}
#| echo: false

set.seed(123)
```

```{r}
# create a vector for gene expression
gene_expr <- sample(x = 1:20, size = 10) # generate 10 random numbers between 1 and 20
gene_expr

# check the structure
class(gene_expr)
str(gene_expr)

# vector length
length(gene_expr)

# show the first/last elements
head(gene_expr)
tail(gene_expr)

```

:::
::: {.column .fragment}

```{r}
# Arithmetic operations
gene_expr + 1
gene_expr - 1
gene_expr * 10
gene_expr / 10
```

::: {.fragment}
```{r}
# Apply some functions
summary(gene_expr)
sum(gene_expr)
mean(gene_expr)
median(gene_expr)
```

:::

:::
:::


# Matrices

A matrix is a **two dimensional** data structure with **rows** and **columns**,
it contains data of the **same data type**.

## Matrices Creation {.smaller}

::: columns
::: {.column}
- Use the `matrix()` function to create a matrix.

```{r}
my_mat1 <- matrix(1:6, nrow = 2)
my_mat2 <- matrix(1:6, nrow = 2, byrow = TRUE)

my_mat1
my_mat2
```

:::

::: {.column .fragment}
- Combine vectors to create matrix.

```{r}
vec1 <- 1:3
vec2 <- 4:6

rbind(vec1, vec2)
cbind(vec1, vec2)
```

<br>

::: {.fragment}
What is the data structure of each row/column of a matrix?
:::

:::
:::


## Matrices Indexing (1) {.smaller}

- Use `[`row_idx`, `column_idx`]` to access element(s).

::: columns
::: {.column}

```{r}
mat <- matrix(1:9, ncol = 3)
mat
dim(mat) # dimensions of the matrix
ncol(mat)
nrow(mat)

mat[1, 2] # element in the 1st row and 2nd column
mat[, 3] # all rows of the 3rd column
```

:::

::: {.column .fragment}

::: {.content-smaller}
Your turns:

- how to get all columns of the 2nd and the 3rd rows?
- how to get the value 5 from the matrix?
:::

::: {.fragment}

```{r}
mat[2:3, ]
mat[2, 2]
```

:::

:::
:::

::: {.notes}
As matrix is vectors extended into 2D, 
if question about drop dim, drop = FALSE
:::

## Matrices Indexing (2) {.smaller}

- Use `[`row_name`, `column_name`]` to access element(s) if names exist.

::: columns
::: {.column}

```{r}
mat

# give names to the columns and rows
named_mat <- mat
colnames(named_mat) <- c("c1", "c2", "c3")
rownames(named_mat) <- c("r1", "r2", "r3")

named_mat

named_mat["r1", ] # all element of the 1st row
```

:::

::: {.column .fragment}

::: {.content-smaller}
Your turns: by using the names of rows and columns,

- how to get 2nd row of the 2nd and the 3rd columns?
- how to get the value 5 from the matrix?
:::

::: {.fragment}
```{r}
named_mat["r2", c("c2", "c3")]
named_mat["r2", "c2"]
```

:::
:::
:::

## Matrices Indexing (3) {.smaller}

::: columns

::: {.column}

- Use logical vector(s) for indexing.

```{r}
mat[c(TRUE, TRUE, FALSE), ]
named_mat[c(TRUE, TRUE, FALSE), c(FALSE, TRUE, TRUE)]
```

:::

::: {.column .fragment}
Your turns:

By using the logical indexing, select the 2nd and 3rd rows, the 1st and 2nd columns of the `named_mat`.

::: {.fragment}
```{r}
named_mat[c(FALSE, TRUE, TRUE), c(TRUE, TRUE, FALSE)]
```

:::

:::
:::

## Matrices Operations {.smaller}

::: columns
::: {.column}

- Check the structure
```{r}
class(mat)
str(mat)
```

::: {.fragment}

- Some maths

```{r}
mat + mat

rowSums(mat)
colSums(mat)
colMeans(mat)
rowMeans(mat)
```

:::
:::

::: {.column .fragment}

- Modify elements


```{r}
mat[1:2, 3] <- c(1, 2)
mat
```

::: {.fragment}

```{r}
mat[1:2, ] <- 10
mat
```

:::

::: {.fragment}
```{r}
#| error: true

mat[1, ] <- c(1, 2)
```

:::

::: {.content-smaller .fragment}
When assigning new values, you must provide either:

- A single value, 
- A vector with the exact number of elements to be replaced, or
- A vector whose length is a factor of the number of elements to be replaced.

:::
:::

:::

# Let's Practice !

## Today's Goals

- Get familiar with data types
- Get familiar with vectors and matrices manipulations
- Simulate your own biological data and test the normality using the Shapiro-Wilk test


