---
title: "Week 8 - Hands-On Examples"
format: 
  html: default
  # pdf: default
date: "2025-03-11"
date-format: " "
categories: 
  - week08
  - exercise
editor_options: 
  chunk_output_type: console
---

The R script is available here:
[link](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/week08/r_w08_exos.R)

## Goals

- Clean, reshape, and manipulate real-world data using {`dplyr`} and {`tidyr`} 
- Integrate tidy data into downstream analysis (e.g., for visualization or statistical modeling)

## Install {tidyverse} and Load the Package

```{r}
# install.packages("tidyverse")
library(tidyverse)
```


## Mini Data Project

A researcher has measured the expression levels of **20 genes** (anonymed as 1 to 20) using the RT-qPCR technique.
The gene expression was assessed in **two structures** of the mouse brain.
Mice ranged in age from 10 to 60 days (**10, 15, 20, 25, 30, 35, 40, 45, 50, 60 days**),
and the experiment was repeated with both **male** and **female** mice,
with **6 animals** (named from A to F) in each group.

According to the researcher, the data was stored in two files, one for each brain structure.
Within each file, rows represent the different ages,
and columns represent the gene, sex, and animal.

A small Gaussian noise has been added to the original data, preserving the overall structure.

The data is available in two CSV files:

- [data_anonym_struc1_noise.csv](../exos_data/data_anonym_struc1_noise.csv)
- [data_anonym_struc2_noise.csv](../exos_data/data_anonym_struc2_noise.csv)

We will focus on the data from the brain structure 1.

1. Please download the `data_anonym_struc1_noise.csv` file.
Observe your data file (for example with a text editor or via terminal):

- Is there a header line? 
- What is the separator between columns?
- Which character was used for decimal points?
- Which character was used for missing data (between two seperators where there's no value)?

2. Import the `data_anonym_struc1_noise.csv` into RStudio, you can use either:

- the `read_csv2()` from the package {`readr`} (`?readr::read_csv2`), or
- use the click-button way and copy-paste the code in your script.

**Don't forget to use/select the appropriate parameters to make sure you import correctly the data.**

Name the data as `data1`.
Convert your imported data to tibble format if it's not the case.

What is the data dimension?

```{r}
data1 <- readr::read_csv2(
  file = "../exos_data/data_anonym_struc1_noise.csv",
  locale = locale(decimal_mark = ",")
)
data1

# or use the read_delim() with appropriate parameters
# data1 <- read_delim(
#   "../exos_data/data_anonym_struc1_noise.csv", 
#   delim = ";", locale = locale(decimal_mark = ",")
# )
```

3. Show the first 10 columns of your data.

```{r}
# classic way
data1[, 1:10]

# use select()
select(data1, 1:10)
```

4. What are the column names?

```{r}
colnames(data1)
```

5. Rename the first column as `age`.

```{r}
data1 <- data1 |>
  rename(age = ...1)

```

6. Reshape data to long format with the `pivot_longer()` function.
What are the columns to be included to pivot into longer format?
 
```{r}
# use column index
data1_long <- data1 |>
  pivot_longer(cols = -1, names_to = "id", values_to = "value")

# use column name
# data1_long <- data1 |>
#   pivot_longer(cols = -age, names_to = "id", values_to = "value")

data1_long 
```

7. Add a column `struc` which contains the name of the measured structure `s1`.

```{r}
data1_long <- data1_long |>
  mutate(struc = "s1")
data1_long
```

8. Extract information about gene, sex and animal from the column `id` using the `extract()` function. Name the new columns as "gene_id", "sex" and "animal".

Hint: Find the patterns for the extraction.
You can use AI to help you to write the regular expression.

```{r}
data1_long <- data1_long |>
  extract(
    col = id,
    into = c("gene_id", "sex", "animal"),
    regex = "([0-9]+)([MF])([A-F])"
  )
```

If you want to write all previous codes together:

```{r}
#| eval: false

data1_long <- readr::read_csv2(
    file = "../exos_data/data_anonym_struc1_noise.csv",
    locale = locale(decimal_mark = ",")
  ) |>
  rename(age = ...1) |>
  pivot_longer(cols = -1, names_to = "id", values_to = "value") |>
  mutate(struc = "s1") |>
  extract(
    col = id,
    into = c("gene_id", "sex", "animal"),
    regex = "([0-9]+)([MF])([A-F])"
  )
```

Now, the data is ready for downstream analysis. Let's focus on gene 1.

9. At age of 10 days, which animal has the highest expression value for gene 1 overall?
And which animal has the highest expression value in each sex?

```{r}
# overall
data1_long |>
  filter(gene_id == "1" & age == 10) |>
  arrange(desc(value)) |>
  slice(1)


# in each sex
data1_long |>
  filter(gene_id == "1" & age == 10) |>
  group_by(sex) |>
  arrange(desc(value)) |>
  slice(1)
```

10. Is there any missing value for gene 1?
If yes, how to remove lines with NA?

```{r}
data1_long |>
  filter(gene_id == "1") |>
  pull(value) |>
  is.na() |>
  table()

data1_long |>
  filter(gene_id == "1") |>
  drop_na(value)
```

11. After removing NAs, how many animals are there for each sex in gene 1?

```{r}
data1_long |>
  filter(gene_id == "1") |>
  drop_na(value) |>
  group_by(sex) |>
  count()
```

12. Summarize the median, mean, and standard deviation of gene 1 expression for both sexes.

```{r}
data1_long |>
  filter(gene_id == "1") |>
  drop_na(value) |>
  group_by(sex) |>
  summarise(
    median_gene1 = median(value),
    mean_gene1 = mean(value),
    sd_gene1 = sd(value)
  )
```

13. Draw a boxplot to show gene 20 expression level (with NA removed) for both sexes at different time point (age).

```{r}
data1_long |>
  filter(gene_id == "1") |>
  drop_na(value) |>
  ggplot(aes(x = as.character(age), y = value)) +
  geom_boxplot(aes(color = sex)) +
  labs(
    x = "Age (days)", y = "Expression Level", color = "Sex",
    title = "Gene 1"
  ) +
  theme_light()
```

At age of 10 days, it seems that there's a difference between male and female at all age.
Let's perform a linear regression to test it.

:::{.callout-note title="Stats Time!"}
**Univariate Linear Regression**

Univariate linear regression is a statistical method used to model the relationship between a single independent variable (predictor, for example the sex) and a dependent variable (response, for example the gene expression). 
The goal is to fit a straight line (linear model) to the data that best represents the relationship.
This can be used for predictions or to understand the strength and direction of the relationship between the variables.
The model is typically represented as:

$$
Y = \beta_0 + \beta_1 X + \epsilon
$$

Where:

- $Y$ is the dependent variable (response),
- $X$ is the independent variable (predictor),
- $\beta_0$ is the intercept,
- $\beta_1$ is the slope,
- $\epsilon$ is the error term.

In R, we use the `lm()` function to perform a linear regression.
:::

```{r}
# fit the model
res_g1_d10 <- lm(
  formula = value ~ sex,
  data = data1_long |> filter(gene_id == "1" & age == 10)
)

# View the summary of model
summary(res_g1_d10)
```

The reference level in sex is "F" (female).

Taking $\alpha = 0.05$ (type I error), coefficient interpretation:

- `Intercept` = 2.1531: on average, the gene expression level for females (the reference group) is 2.1531. The p-value indicates that the baseline gene expression for females is significantly different from zero.
- `sexM` = -1.0233: on average, gene expression for males is 1.0233 units lower than for females, assuming all other factors remain constant. The p-value indicates that sex does have a significant effect on gene expression.

Overall, the linear regression suggest that there is a statistically significant difference in gene expression between males and females at age of 10 days for gene 1.
Males have lower expression levels than females, with the difference being approximately 1.0233 units.


## Bonus

Use the `read.table()` function to import the data and continue to reshape the data based on the imported data.

(Check the approporiate parameters to be included with `?read.table`)

```{r}
#| code-fold: true
#| code-summary: "Correction"

# import data with the basic function
data1 <- read.table(
  file = "../exos_data/data_anonym_struc1_noise.csv",
  header = TRUE, sep = ";", dec = ",", na.strings = ""
)
# convert to a tibble
data1 <- as_tibble(data1)

# transform to long format and 
data_long <- data1 |>
  rename(age = X) |>
  pivot_longer(cols = -1, names_to = "id", values_to = "value") |>
  mutate(
    id = sub("X", "", id), # remove the X in ID
    struc = paste0("s", 1) # store info of brain structure
  ) |>
  extract( # extract column into multiple columns
    id,
    into = c("gene_id", "sex", "animal"),
    regex = "([0-9]+)([MF])([A-F])"
  ) |>
  mutate(
    gene_id = paste0("g", gene_id) # recode the gene column
  )

data_long
```

---

#### Good job! 👏👏 You've made great progress in mastering data manipulation techniques.


