---
title: "Week 4 - Hands-On Examples"
format: 
  html: default
  # pdf: default
categories: 
  - week04
  - exercise
editor_options: 
  chunk_output_type: console
---

The R script is available here:
[link](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/week04/r_w04_exos.R)

## Goals

- Understand and use operators to filter data with precision.
- Understand function's basic structure, learn to use new functions, write functions to automate tasks.

## Import Data

A gene-level diffenrential expression (DE) analysis was performed to compare SET1 samples to WT samples using data from `read-counts.csv`.

The analysis result is available via this [link](https://raw.githubusercontent.com/InforBio/IOC/refs/heads/main/ioc_r/exos_data/toy_DEanalysis.csv).

- Please donwload the result file and upload it to your data folder.
- Import the data using `read.csv()` function.
(See the documentation with `?read.csv`) 
Name the imported results `de_res`.

```{r}
getwd()
```

```{r}
de_res <- read.csv(
  file = "../exos_data/toy_DEanalysis.csv",  # replace the path with your own
  header = TRUE
)
```


## Exercises

1. Check the structure of `de_res` using an appropriate R function. What are the dimensions?

```{r}
str(de_res)
```

The result is a data frame with `r nrow(de_res)` rows and `r ncol(de_res)` columns.


The result contains following columns:

- `gene_name`: gene name
- `baseMean`: mean of normalized counts for all samples
- `log2FoldChange`: log2 fold change
- `lfcSE`: standard error
- `stat`: Wald statistic
- `pvalue`: Wald test p-value
- `padj`: adjusted p-values (Benjamini-Hochberg procedure)

::: {.callout-tip title="Stats Time!"}
**Multiple Tests Correction**

- Why multiple testing is a problem?

If we do 100 tests simultaneously and set and use significance level at 0.05,
$$
\begin{aligned}
P(\text{at least 1 significant result by change}) &= 1- P(\text{non significant results}) \\
&= 1 – (1 - 0.05)^{100} \\
&= 0.99
\end{aligned}
$$

- Multiple test correction
<ul>
  <li> **Bonferroni** correction: control the significance level.</li>
  <p>
$$
\begin{aligned}
P(\text{at least 1 significant result by change}) &= 1 – (1 - \frac{0.05}{100})^{100} \\
&= 0.049
\end{aligned}
$$
  </p>
  <li> **FDR** (False discovery rate): control the proportion of false positive amongst all significant results, *e.g.*: Benjamini-Hochberg (BH) procedure.</li>
</ul>

:::

2. Extract results for these genes: *RNR1*, *PIR3*, *SRP68*.

```{r}
de_res[de_res$gene_name %in% c("RNR1", "PIR3", "SRP68"), ]
```

3. Order the result by decreasing p-value using the `order()` function.
(See `?order()`)

Hints:

- `order()` can return the indices of the vector elements in ascending or descending order
- Apply `order()` to the desired column
- Use the returned indices to rearrange the dataframe

```{r}
## fast check how order() works
order(c(0.2, 0.5, 0.3))

## arrange the data frame
de_res <- de_res[order(de_res$pvalue), ]
```

4. Use `ifelse()` to categorize genes. 
Add a new column, `gene_category`, that assigns categories:

- "up" if `log2FoldChange > 0.5`.
- "down" if `log2FoldChange < -0.5`.
- "neutral" otherwise.

```{r}
de_res[["gene_category"]] <- ifelse(
  test = de_res[["log2FoldChange"]] > 0.5,
  yes = "up",
  no = ifelse(
    test = de_res[["log2FoldChange"]] < -0.5,
    yes = "down",
    no = "neutral"
  )
)
```

5. Use `table()` to count the occurrences of each gene category.

```{r}
table(de_res[["gene_category"]])
```


6. Write the code to create a subset of significant upregulated genes with:

- Adjusted p-value (`padj`) < 0.05.
- Log2 fold change (`log2FoldChange`) > 1.

```{r}
up_regulated <- de_res[
  de_res$padj < 0.05 & de_res$log2FoldChange > 1,
]
up_regulated
```

7. Write the code to create a subset for genes that meet any of the following conditions:

- `pvalue < 0.01` and `log2FoldChange > 0.5`.
- `pvalue < 0.01` and `log2FoldChange < -0.5`.

```{r}
target_gene <- de_res[
  de_res$pvalue < 0.01 &
  (de_res$log2FoldChange > 0.5 | de_res$log2FoldChange < -0.5),
]
target_gene

# or test the absolute value of log2FoldChange
de_res[
  de_res$pvalue < 0.01 & abs(de_res$log2FoldChange) > 0.5,
]

```

8. Write a function to automate "de_res" filtering for genes
with a p-value less than or equal to a custom cutoff.

```{r}
filter_p <- function(cutoff = 0.05) {
  de_res[de_res$pvalue <= cutoff, ]
}
filter_p(0.01)
filter_p(0.0001)
```

9. Based on the function created in question 8,
modify the function to allow output ordered by any desired column in `de_res`.

Hints: You need an extra parameter to specify the wanted column and
another parameter to fix the cutoff.

```{r}
filter_col <- function(col = "pvalue", cutoff = 0.05) {
  de_res[de_res[[col]] <= cutoff, ]
}
filter_col(cutoff = 0.0001)
filter_col(col = "log2FoldChange", cutoff = -1)
```

10. Based on the function created in question 9,
add a step in the body of the function to extract
the gene names that met the filter condition.

```{r}
extract_gene <- function(col = "pvalue", cutoff = 0.05) {
  de_res[de_res[[col]] <= cutoff, ][["gene_name"]]
}
extract_gene(cutoff = 0.0001)
extract_gene(col = "log2FoldChange", cutoff = -1)
```

---

#### Bravo! 🎉 You've learned the basics of R, and you're already making great progress, keep it up!


